---
title: "lab_02"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? 
**A1.**Atmospheric and Oceanic Science offered the most amount of courses with Climate in the title.

```{r}
climate_courses <- umd_courses |>
  filter(str_detect(title, "Climate")) |>
  filter(seats >= 1) 
```

```{r}
climate_courses_2 <- climate_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    max_seats = max(seats),
    min_seats = min(seats),
  ) |>
  arrange(desc(max_seats), desc(min_seats))
```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why? 
**A2.** The departments that are the most interesting to me are Finance and Government and Politics. I dont know much about finance, but I don't think that they would talk much about climate change. As for Goverment and Politics, they may be referring to the political climate but it is still interesting. 

**Q3** Which departments offered the most courses in the Fall 2024 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?
**A3.**The departments that offered the most courses in the Fall 2024 term are Business and Management with 174, School of Music with 158, and Information Studies with 137. The departments in the same term that have at least one syllabus posted are under the column, "courses_with_syllabus." There are many listed but a few examples are Business and Management with 106 and Economics with 54. The top five departments are Business and Management, School of Music, Information Studies, English, and Teaching and Learning, Policy and Leadership in terms of courses. The interesting part is that while there may be many courses, not all post their syllabi. For example, Business and Management had the most courses with 174 but only posted 106 syllabi. Or English had 125 courses but only posted 11 syllabi.

```{r}
fall2024_courses <- umd_courses |>
  filter(term == 202408) |>
  group_by(department) |>
  summarise(
    courses_department = n(),
    syllabi = sum(syllabus_count > 0, na.rm = TRUE), 
  )|>
   filter(syllabi > 0) |>  
arrange(desc(courses_department), desc(syllabi))

head(fall2024_courses)
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name). Write a sentence describing the pattern of spending over time. 
**A4.** Wes Moore's committee spending goes up each month, starting from 6000.3 in 2021-05-03 to	3121109.8 in 2022-10-01. The spending patterns stay around the same amount from the months of 06-2021 to 12-2021 and jump with the start of 2022. 

```{r}
maryland_expenses <- maryland_expenses |>
mutate(expenditure_date = mdy(expenditure_date)) 
```

```{r}
glimpse(maryland_expenses)
```
```{r}
wes_moore <- maryland_expenses|>
filter(str_detect(committee_name, "Moore")) |>
mutate(month = floor_date(expenditure_date, "month")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount, na.rm = TRUE)) |>
  arrange(desc(total_amount))

wes_moore
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).
**A5**
In-state was 77,735,787 and Out of state was 51,552,960

```{r}
maryland_expenses <- maryland_expenses |>
  mutate(
    spending_location = case_when(
      str_detect(address, "Maryland") ~ "In-state",
      TRUE ~ "Out of state"
    )
  )
```

```{r}
maryland_expenses |>
  group_by(spending_location) |>
  summarise(total_amount = sum(amount, na.rm = TRUE))
```


**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. **A6.**
My question was one that I did not ask in the pre_lab, but it is Which committees spent more than $1,000,000 on individual categories of expenses and how much did they spend?

These were my results: 
Moore Wes For Maryland- Transfers Out to Other Maryland Treasurers- 4987002	
Franchot Peter Friends Of- Transfers Out to Other Maryland Treasurers- 3649235	
Moore Wes For Maryland- Salaries and Other compensation- 3428264	
Blair David for Montgomery- Media- 3121288	
Lierman Brooke for Maryland- Media- 2706907	
Adams Tim(Timothy) Friends of- Media- 2404310	
Perez Tom for Maryland- Transfers Out to Other Maryland Treasurers- 2279378	
Blair David for Montgomery- Salaries and Other compensation- 2173884	
Miller Aruna Friends of- Transfers Out to Other Maryland Treasurers- 2010449	
Haire Jessica Friends of- Media- 1778237

```{r}
committee_category_totals <- maryland_expenses |>
  group_by(committee_name, expense_category) |>
  summarise(
    total_spent = sum(amount, na.rm = TRUE),
    .groups = "drop"   # Ungroup after summarising
  )
```

```{r}
high_spending_categories <- committee_category_totals |>
  filter(total_spent > 1000000) |>
  arrange(desc(total_spent))
```

```{r}
high_spending_categories
```
