---
title: "Reverse Engineering Project"
author: Ellie Jornlin
date: Oct. 18th 2025
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [Nearly 250 women have been fatally shot by police since 2015] (https://www.washingtonpost.com/graphics/2020/investigations/police-shootings-women/)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(stringr)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
fatal_police_shooting <- read_csv("data/fatal-police-shootings-data.csv") 
fatal_police_shooting_agencies <- read_csv("data/fatal-police-shootings-agencies.csv")

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 


```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: The Post found only one other fatal shooting that closely matched Taylor’s case — a Black woman, unarmed, killed during a raid at home while a boyfriend shot at police. But 139 other cases shared one or more of the circumstances in which Taylor was killed. Of the 247 women fatally shot, 48 were Black and seven of those were unarmed.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

When filtered for women, regardless of race, the total of victims fatally shot is 422, nearly double the amount reported by the Post (not sure how to get nearly 250). Of the 422 women shot by the police, 87 were Black and female while nine were Black, female, and unarmed; Breanna Taylor is part of this list. I am unable to verify if these women were at home when they died because their addresses are not included in the data set. Is there a way to generate a change where the longitute and latitude data into locations? Or would we need to look each one up and create a new column for it?

```{r}
fatal_race_armed <- fatal_police_shooting |>
 filter(!is.na(race) & !is.na(armed_with))
```

```{r}
# Put code to reverse engineer sentence here
fatal_race_armed |>
select(race, gender, armed_with, name) |>
  filter(
    race == "B",
    gender == "female"
  )
# Display results of code below this codeblock
```

```{r}
fatal_police_shooting |>
select(race, gender, armed_with, name) |>
filter(str_detect(race, "B")) |>
filter(str_detect(gender, "female")) |>
  filter(str_detect(armed_with, "unarmed"))
```

```{r}
fatal_race_armed |>
select(race, gender, armed_with, name) |>
  filter(
    gender == "female"
  ) 
```

### Sentence 2

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 3

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```


### Sentence 4

* **Sentence text**: As with fatal police shootings of men, the vast majority of the women killed were armed with a potential weapon at the time, although slightly less often...[and]... About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed.

* **Analysis summary**: The first finding is that the majority of women killed were armed with a potential weapon at the time. We found that 90.5% of women killed were armed vs. 9.5% unarmed. For men, 94.3% were armed vs. 5.7% unarmed. This is accurate when compared to the article. However, the article lists that 89 percent of the women were armed, compared to 91 percent of the men. The percentages are a bit off, but that is due to how they classified armed vs. unarmed. The second finding is that about 31%, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22% of the 5,362 men killed. We got 31% of women killed had mental health issues and 19.2% of men had mental health issues. The percentage for women is correct, but we are unsure how to get the exact 22% of men. It may be because the data spans further than what the Post used. 

```{r}
# Put code to reverse engineer sentence here
library(dplyr)


armed_binary <- fatal_police_shooting |>

  filter(!armed_with %in% c("unknown", "undetermined") & !is.na(armed_with)) |>
  mutate(
    armed_status = case_when(
      armed_with == "unarmed" ~ "unarmed",
      TRUE ~ "armed" 
    )
  ) |>
  group_by(gender, armed_status) |>
  summarise(
    n = n(),
    .groups = "drop"
  ) |>
  group_by(gender) |>
  mutate(
    total = sum(n),
    percent = round(100 * n / total, 1)
  )

armed_binary
```
```{r}
mental_health_by_gender <- fatal_police_shooting |>
  mutate(gender = tolower(trimws(gender))) |>
  filter(gender %in% c("male", "female")) |>
  group_by(gender) |>
  summarise(
    total = n(),
    mental_health_cases = sum(was_mental_illness_related == TRUE, na.rm = TRUE),
    percent_mental_health = round(100 * mental_health_cases / total, 1),
    .groups = "drop"
  )

mental_health_by_gender
```

### Sentence 5

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-