---
title: "lab_11"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries for working with data, plus rvest and lubridate.

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.
```{r}
library(tidyverse)
library(janitor)
library(rvest)
library(lubridate)
library(stringr)


options(scipen = 999)
```


Let's get to scraping.

## Questions

**Q1**. Scrape the listing of available Maryland physician disciplinary alerts at <https://www.mbp.state.md.us/disciplinary_2024.aspx> into a dataframe. You should have three columns, one of which is a date, so make sure the date column has a date datatype. What's the most common sanction, and how many times is it mentioned in the 2024 alerts?

**A1** 
The most common sanction is Summary Suspension Affirmed and Surrender of License, which are both mentioned 7 times. 

```{r}
sanctions_2024 <- "https://www.mbp.state.md.us/disciplinary_2024.aspx" |>
  read_html() |>
  html_table() |>
  as.data.frame() |>
  clean_names() |>
  mutate(date = mdy(date))

sanctions_2024 |>
  group_by(sanction) |>
  summarise(
    count = n(),
    date = first(date)  
  ) |>
  arrange(desc(count))
```

**Q2** Next, let's scrape the list of press releases from Maryland's Office of the Public Defender, <https://www.opd.state.md.us/press-releases>. This isn't a table, so you'll need to use `html_elements()` and your browser's inspector and do some clean up on the results. The result should be a dataframe with two columns that contain the date and title, and the date column should have a date datatype. The challenge here is figuring out how to isolate the releases.

When you finish scraping into a dataframe, write code to find the press releases that have the word "police" in the title.

**A2** 
There are 9 press releases that have the word "police" in the title. They are: 

In Aftermath of Video Depicting Excessive Police Force, Public Defender Calls for the Implementation of Body Worn Cameras by Ocean City Police
Public Defenders Testify About Police Violence and Misconduct Statewide; Need for Police Accountability, Transparency, and Reform
Public Defender Seeks to Identify Cases Potentially Impacted by Breach of Salisbury Police Department
Statement of Kirsten Getty-Downs, District Public Defender for Baltimore City, Regarding Police Corruption By Former BPD Officer Matthew Ryckman
Another Problematic Police Body-Worn Camera Video Identified; Baltimore State’s Attorney Still Failing to Disclose Information in Impacted Cases
Public Defender Discloses Video Evidence of Officers Planting Drugs; Demands Swift Action by Prosecutors and Thorough Review by Police
OPD Statement on the Baltimore Police Department’s Secret Aerial Surveillance Program
OPD Applauds DOJ Report, Urges Greater Transparency by the Baltimore City Police Department
Police Continue to Illegally Arrest Children in Baltimore

```{r}
opd_pressrelease <- read_html("https://opd.state.md.us/press-releases")

press_release <- opd_pressrelease |>
  html_elements('p') |>
  html_text() |>
  as_tibble() |>
  separate(value, c("date", "title"), sep = ":", extra = "merge")
```

```{r}
pressrelease_clean <- press_release |>
  filter(date != "", !is.na(title)) |>
  mutate(
    date = mdy(str_trim(date)),
    title = str_squish(title)
  )
```

```{r}
police_releases <- pressrelease_clean |>
  filter(str_detect(title, regex("police", ignore_case = TRUE))) |>
  select(date, title)

police_releases
```

**Q3** Sen. Chris Van Hollen, D-Maryland, has posted press releases at <https://www.vanhollen.senate.gov/news/press-releases>. It would be great to have them in a dataframe that has the following columns: date, title and url.

To do this, you will need to scrape the page's html and save that to a variable, and *then* extract the dates, titles and urls into separate dataframes using html_elements(). There are two different elements that contain the press release URL; pick one of them, and you may need to invoke an HTML `class` value to do it. The function `html_text()` pulls out the contents of a tag, but for urls we want the HTML attribute. Rvest gives you a way to extract the URL from a link; google to find out what it is. Remember how we turn a list into a dataframe.

At the end, you'll have three dataframes that you want to combine into a single dataframe. When we want to combine the rows of identical dataframes, we used `bind_rows()`. If you were combining columns instead of rows, there's a similar function. Find out what it is and use it to put all of the dataframes together into a single one.

When you're done, rename the columns so they make sense, then make sure the date column is an actual date.

Finally, tell me what questions you could ask of this data if you had all of his press releases. Be creative.

**A3**
How many press releases does Van Hollen release per year? Has the number grown or decreased?
What are the most common words or phrases in the titles of the press releases?
Which year has the most press releases? Are there particulars seasons or months that have more?
Which day of the week are press releases most commonly published?
Are there specific topics of press release that are more common? 
Are press releases more common around major political events?

```{r}
press_page <- read_html("https://www.vanhollen.senate.gov/news/press-releases")
```

```{r}
press_dates <- press_page |>
  html_elements("p") |>
  html_text() |>
  str_trim() |>
  as_tibble() |>
  rename(raw_date = value) |>
  mutate(
    cleaned_date = str_replace_all(raw_date, "\\.", "/"),  
    date = mdy(cleaned_date)
  ) |>
  select(date)
```

```{r}
press_titles <- press_page |>
  html_elements("h2") |>
  html_text() |>
  str_trim() |>
  as_tibble() |>
  rename(title = value) |>
  filter(!title %in% c("Filter Results", "website-search"))  
```

```{r}
press_urls <- press_page |>
  html_elements("a") |>
  html_attr("href") |>
  as_tibble() |>
  rename(url = value) |>
  filter(
    str_starts(url, "https://www.vanhollen.senate.gov/news/press-releases/"),
    !str_detect(url, "\\.(png|jpg|jpeg|gif|svg|ico)$"), 
    !str_detect(url, "^/assets/"),                      
    url != "/"
  ) |>
  distinct()  
```

```{r}
press_releases <- bind_cols(press_titles, press_dates, press_urls)
```

```{r}
head(press_releases)
```




