---
title: "lab_06"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# load your libraries
library(tidyverse)
library(lubridate)
library(janitor)
library(readr)
library(dplyr)
```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. We'll use some more 911 overdose call data to do this. The first thing we want to do is to combine multiple counties' data into a single dataframe so that we can ask some questions. First, let's combine data from Cecil, Carroll and Allegany counties into a new dataframe.

*Before* you combine them you'll need to clean up some of the column names to make the data easier to work with - make sure you have loaded the library to do that. You also need to make sure that each column has the same name and datatype (you can check the datatypes using `glimpse`). If any of them does not, you need to fix that.

```{r}
# load and combine the call data from those three counties
allegany_911 <- read_csv("data/allegany_911.csv")
carroll_911 <- read_csv("data/carroll_911.csv")
cecil_911 <- read_csv("data/cecil_911.csv") 

glimpse(carroll_911)
glimpse(allegany_911)
glimpse(cecil_911)
```

```{r}
carroll_911 <- carroll_911 |>
  rename(
    date = Date,
    location = Location
  )
```
```{r}
carroll_911
```
```{r}
combined_911 <- bind_rows(allegany_911, carroll_911, cecil_911)
```

Then we'll join some demographic data with Baltimore City 911 calls and ask some questions of that. For that, you'll need to load the Baltimore City 911 data and the ZCTA demographic data from the data folder, and join them based on the zip code. Remember, the zip codes in both dataframes must be the same datatype (and should be characters). You should start with your Baltimore City 911 dataframe in creating your join.

```{r}
baltimore_911 <- read_csv("data/baltimore_911.csv")
md_zcta <- read_csv("data/maryland_zcta.csv")

glimpse(baltimore_911)
glimpse(md_zcta)
```
```{r}
md_zcta <- md_zcta |> mutate(across(ZCTA5N, as.character)) |>
rename(zip_code = ZCTA5N) |>
clean_names()

baltimore_911 <- baltimore_911 |> mutate(across(zip_code, as.character))

glimpse(baltimore_911)
glimpse(md_zcta)
```


```{r}
combined_balty_calls <- md_zcta |>
right_join(baltimore_911)

combined_balty_calls
```

## Answer questions

Q1. Let's start with our data from Allegany, Cecil and Carroll counties. Write code to generate a dataframe that shows the total number of calls for each county. What's the order of counties from most calls to least?

A1.Carroll Counting had the highest number of calls with 448, Cecil had the second highest with 438 and Allegany had the least number of calls with 418.

```{r}
combined_911 |>
   group_by(county) |>
  summarise(count = n()) |>
 arrange(desc(count))
```

Q2. In your combined dataframe that you used in A1, add a column for the month of each call, then write code to show the total number of calls per county and month. Are there any outliers or noteworthy results? Describe the general pattern of the results.

Then calculate the most calls per county and location. Which location had the most calls - where and what is that location?

A2. 
1. There are not any noteworthy outliers, but the general pattern is that the number of calls is higher in the warmer months than the colder months. For example in Allegany county, the months with the highest number of calls are August, September, May and July while the lowest are December, January and November. Also in Carroll the highest is August and the lowest is September. In Cecil the highest is August and the lowest is April, which may be an outlier but the second lowest is December. 
2. The location that had the highest number of calls is 314 Grove Neck Road in Earlville, MD which is in Cecil County. 


```{r}
combined_911 <- combined_911 |>
  mutate(
    date = as.Date(date), 
    month = month(date, label = TRUE, abbr = TRUE)
  )
```

```{r}
county_per_month <- combined_911 |>
  group_by(county, month) |>
  summarise(total_calls = n(), .groups = "drop") |>
  arrange(county, desc(total_calls))

county_per_month
```
```{r}
county_per_location <- combined_911 |>
  group_by(county, location) |>
  summarise(total_calls = n(), .groups = "drop") |>
  arrange(desc(total_calls))
  
county_per_location
```

Q3. Let's turn to the Baltimore City data. You're interested in exploring calls in zip codes with a high percentage of children under 18. Write code to isolate calls in zip codes where the percentage of under 18 population is at least 75% and then show the zip code and population and how many calls occurred in each of those zip codes.

Where are those zip codes in the city?

Add a column to your code that calculates the number of calls per 1,000 people for those zip codes. To calculate a per capita rate, you might find this short guide handy: [https://observablehq.com/\@palewire/per-capita-calculator](https://observablehq.com/@palewire/per-capita-calculator){.uri}.

Which zip code has the highest rate of calls per 1,000 people? Find the neighborhoods in that zip code that are listed in the data - you can use group_by or distinct to do this. What are some of those neighborhoods, and what else can you tell me about the population there?

A3.
1. The zip codes with the highest amount of calls are 21217	with a population of 37111 and 774 calls. Next is 21223	with a population of 26366 and 525 calls. Finally is 21202 with a population of 22832 and 325 calls. 
2. The zip code with the highest rate of calls per 1,000 people is 21217. The neighboroods listed in that area are Penn North, Sandtown-Winchester, Upton, Druid Heights, Parkview/Woodbrook, Reservoir Hill, Mondawmin, Harlem Park, Easterwood, Madison Park, Bolton Hill, Midtown-Edmondson, Druid Hill Park, Jones Falls Area, Mid-Town Belveder and Bridgeview/Greenlawn. The neighborhood with the highest amount of calls is Penn North with 177, next is Sandtown-Winchester with 174 and finally Upton with 163. The percentage of people number 18 is pretty high, with 81.5%. The median age is 34.3. The percentage of population over 65 is 11.1%. 

```{r}
glimpse(combined_balty_calls)
```

```{r}
pct_underage_zips <- combined_balt_calls |>
  filter(pct_under_18 >= 75) |>
  group_by(.data$zip_code, .data$population) |>
  summarise(total_calls = n(), .groups = "drop") |>
  arrange(desc(total_calls)) |>

mutate(pct_per_thousand = total_calls / population * 1000)|>
  arrange(desc(population))
  
pct_underage_zips
```

```{r}
combined_balty_calls |>
  filter(zip_code == "21217") |>
  group_by(neighborhood, pct_under_18, pct_pop_65, medage) |>
  summarise(count_total = n(), .groups = "drop") |>
  arrange(desc(count_total))
```

Q4. Choose your own adventure: I want you to come up with a question you could ask of the Baltimore City data that combines both the calls and the demographic information. You must use both demographics and the overdose call data in your work.

A4:
The question that I asked is for the zip codes with the highest number of calls, what is the comparison between renters and homeowners?

The top 3 zip codes with the most amount of calls are 21217, 21223 and 21215. The renter and homeowner comparisons are already in percentages which is helpful, but overall the zip codes with higher renters receive more calls compared to homeowners. In the first zip code there were 774 calls with 70% renters and only 21.7% homeowners. In the second there were 525 calls with 64.1% renters and 22.5% homeowners. The third most amount of calls was 500 with 49.7% renters and 35.7% homeowners. The 3 zip codes with the least amount of calls are 21237 with 4 calls, 35.3% renters and 48.9% homeowners. The next lowest is 21208 with 3 calls, 32.8% renters and 46.9% homeowners. The zip code with the least amount of calls is 21228 with 1 call and there are 28.7% renters and 52.8% homeowners.

```{r}
owners_vs_renters <- combined_balty_calls |>
  group_by(zip_code, pct_rent, pct_homeowner, population) |>
  summarise(total_calls = n(), .groups = "drop") |>
  arrange(desc(total_calls))

owners_vs_renters
```
