---
title: "lab_05"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse, plus any other packages you will need to clean data and work with dates.
library(tidyverse)
library(janitor)
library(lubridate)
library(stringr)
```

## Get Our PDF

We'll be working with the [911 overdose calls from Baltimore County](https://drive.google.com/file/d/1qkYuojGF_6WKFr5aNQxmewDzcKyOiJFr/view?usp=share_link). You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to extract the tables within it, export that to a CSV file, load it into RStudio and ask some questions.

## Extract Data from PDF Using Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds or longer.

This PDF has a single table spread over multiple pages to extract. We're going to make a single dataframe from this table, exporting it to a CSV file that you will load into R. In Tabula, highlight the table and click the "Preview & Export Extracted Data" button. You may want to play with including or excluding the column headers - YOU SHOULD HAVE FIVE COLUMNS OF DATA.

Save the CSV (it should be called `tabula-Baltimore County; Carey, Samantha log OD.csv` by default) to your lab_05/data folder.

From there, you will need to read in the data, and add or fix headers if necessary. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)).

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers should not contain spaces and they should have meaningful names, not "x1" or something similar. How you do that is up to you, but you can use select() with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns. Importantly, you'll need to ensure that any columns containing a date actually have a date datatype. Our friend `lubridate` can help with this.

```{r}
baltcounty_overdoses <- read_csv("tabula-Baltimore County; Carey, Samantha log OD.csv") |> clean_names()

baltcounty_overdoses
```
```{r}
clean_baltcounty_overdoses <- baltcounty_overdoses |>
  rename(
    date = dat_format,
    event_type = evtyp,
    location = loc
  ) |>
  mutate(date = mdy(date)) |>
  separate(location, into = c("event_type_from_loc", "street_address"), sep = ",", fill = "right", extra = "merge") |>
  mutate(
    street_address = if_else(is.na(street_address), event_type_from_loc, street_address),
    event_type = str_squish(coalesce(event_type, event_type_from_loc)),  # 
    street_address = street_address |>
      str_to_upper() |>
      str_replace_all("&", "AND") |>
      str_remove("\\(.*?\\)") |>
      str_remove("\\*.*?\\*") |>
      str_remove(";.*") |>
      str_squish()
  ) |>
  select(-event_type_from_loc) 

clean_baltcounty_overdoses
```

```{r}
clean_baltcounty_overdoses <- clean_baltcounty_overdoses |>
  separate(dat_formcaatse_nbr, into = c("time", "case_number"), sep = " ")

clean_baltcounty_overdoses
```

## Answer questions

Q1. Write code to generate the number of calls that occurred on each date. Which date in 2022 had the most overdose calls, and how many? Look at the total number of rows in your result and explore the range of dates - based on your result, do you believe there are any days with no overdose calls at all? Explain why or why not.

A1.
July 14, 2022 had the most amount of calls with 23, tied with October 4, 2022. I do not believe that there are dates with no calls as I arranged the data by descending order and even the date with the lowest amount of calls has one. All dates are positive numbers as well. 

```{r}
calls_per_date <- clean_baltcounty_overdoses |>
  group_by(date) |>
  summarise(count = n()) |>
 arrange(desc(count)) 

calls_per_date
```

Q2. You want to understand if there's a pattern in the day of the week that overdose calls are made. Add a column to your dataframe that displays what day of the week each date represents. You should search for how to do that using lubridate. Then write code to calculate the number of calls for each day of the week, and add a column to that result that calculates the percentage of all calls that occurred on each day of the week (so you want a dataframe with the day of the week, total number of calls and the percentage of calls on that day out of the total number of all calls). Describe your findings to me.

A2.
Saturday had the highest number of total calls with 620 (15.2%) and Thursday had the lowest number of total calls with 520 (12.8%). The top 3 days with the highest number of calls are Saturday, Sunday and Friday which is not surprsing. It would make sense that more calls occur on the weekend as people tend to use more substances on the weekend. 

```{r}
clean_baltcounty_overdoses <- clean_baltcounty_overdoses |>
  mutate(
    day_of_week = wday(date, label = TRUE, abbr = FALSE) 
  )
```

```{r}
calls_by_day <- clean_baltcounty_overdoses |>
  group_by(day_of_week) |>
  summarise(total_calls = n()) |>
  mutate(percent_of_total = round((total_calls / sum(total_calls)) * 100, 1)) |>
  arrange(desc(total_calls))
calls_by_day
```

Q3. Now let's look at locations. Which ones have the most calls? How would you describe them (feel free to search for more information on them)? Is there anything about the structure of the original data that might make you less confident in the counts by location or date?

A3.
I needed to get some help from the internet to change my cleaning process in question 1 (I think my tabula data combined some of the event types with the street addresses so it was giving me confusing data. I figured it out and now have clean addresses.) Anyway, the location with the most calls is 4540 Silver Spring Rd with 36 calls. This address brings me to a 3 bed, 2 bath house that is 2,340 sqft. Zillow says that it was sold on December 21, 2023. The second most amount of calls is Royal Farms with 17 calls. The data could be a bit confusing at the original dataset has MCDONALDS; 7500 PULASKI HY but then it also has stores without their addresses (i.e. Royal Farms).

```{r}
calls_by_location <- clean_baltcounty_overdoses |>
  group_by(street_address) |>
  summarise(total_calls = n()) |>
  arrange(desc(total_calls))

calls_by_location
```

Q4. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A4.
One story idea I have is the number of overdoses that occur at homes or residences versus stores such as Walmart, Wawa and Royal Farms. Another idea is to look into the times that these incidents occur, not just the days of the week. 
