---
title: "lab_10"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
library(tidyverse)
library(sf)
library(janitor)
library(tigris)
library(tidycensus)
```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Maryland voters made history last November by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown *based on the total of their respective contests*. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1.**
This map demonstrates that the counties in which Moore's percentage is leading by the highest amount are Carroll, Kent and Queen Anne's. He leads by a moderate percentage in Harford, Talbot and Dorchester. It continues, but the lowest are Baltimore City and Prince George's. Close behind are Charles and Montgomery. This illustrates that Moore's strongest relative performance came in more rural or swing counties where Brown remained competitive in more urban and heavily Democratic counties. 

```{r}
md_votes <- read_csv("data/md_statewide_county.csv")
```
```{r}
md_votes <- md_votes |>
  mutate(
    GovernorTotal = Cox + Moore + Lashar + Wallace + Harding + Sefcik + GovWritein,
    AGTotal  = Peroutka + Brown + AgWritein,
    MoorePct = (Moore / GovernorTotal) * 100,
    BrownPct = (Brown / AGTotal) * 100,
    DiffPct = MoorePct - BrownPct,
    County = tolower(County)
  )
```

```{r}
md_counties <- counties(state = "MD", cb = TRUE, year = 2020) |>
  st_as_sf() |>
  rename(County = NAME)
```
```{r}
md_votes <- md_votes |>
  mutate(GEOID = as.character(GEOID))
```

```{r}
md_map <- md_counties |>
  left_join(md_votes, by = "GEOID")
```

```{r}
ggplot(md_map) +
  geom_sf(aes(fill = DiffPct)) +
  scale_fill_continuous(type = "viridis") +
  theme_minimal()
```
**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland from last March. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.**
I turned the color scale to log so that I could see the differences in data more easily. However, I was a bit confused at what the numbers meant so I turned back to my original dataset as well. Most counties have moderate to low outage rates, as it was just one month's worth of data, but there are a few outliers. Talbot County haas an outage rate of 1989 per 10,000 customers which is very high compared to other counties which might indicate a major event in Talbot during the month of March. Some counties have very low rates, like Frederick and Washington. These counties are close together as well showing that Western counties largely were less affected than Eastern and Southern counties. If I had access to the data of other months, I would be able to have a better understanding of patterns and trends of outages in Maryland counties. Only having the data from one month makes it difficult to determine the quality of each county's power capabilities or weather trends. However, it does allow me to determine if there may have been specific events during March that caused or did not cause outages depending on the area. 

```{r}
outages <- read_csv("data/county_outages_march2023.csv") |> clean_names()

glimpse(outages)
```
```{r}
outages <- read_csv("data/county_outages_march2023.csv") |>
  rename(County = county) |>         
  mutate(
    County = str_to_upper(County),
    outage_rate = (outage_customers / total_customers) * 10000
  )
```
```{r}
md_counties <- md_counties |>
  mutate(County = str_to_upper(str_squish(County)))
outages <- outages |>
  mutate(County = str_to_upper(str_squish(County)))
```

```{r}
md_outage_map <- md_counties |>
  left_join(outages, by = "County")
```

```{r}
ggplot(md_outage_map) +
  geom_sf(aes(fill = outage_rate)) +
  theme_minimal() +
  scale_fill_viridis_c(option = "magma", trans = "log", na.value = "grey50")
```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change. What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.**
A major county that stands out to me is Caroline county which clearly has the highest amount of electric vehicle prevalance, or at least where the number of EV's grew the most and had the highest percentage growth. The lowest counties are Washington, Kent, Dorchester and a few others. 

The counties that stand out now are Howard, Montgomery, Frederick and Anne Arundel. The first map highlighted Caroline County as it had the largest absolute growth regardless of population. The second map emphasizes where EVs are most concentrated among residents, rather than just raw growth. This shows that some of the more populated and most likely wealthier counties have a higher density of EVs, even if their total growth isn't the biggest. 

```{r}
ev_data <- read_csv("data/ev_2020_2023.csv") |> 
  rename(County = county)
```
```{r}
md_counties <- md_counties |>
  mutate(County = str_to_title(str_trim(County)))

ev_data <- ev_data |>
  mutate(County = str_to_title(str_trim(County)))
```

```{r}
md_ev_map <- md_counties |>
  left_join(ev_data, by = "County")
```

```{r}
md_ev_map <- md_ev_map |>
  mutate(EV_Pct_Change = (march_2023 - july_2020) / july_2020 * 100)
```

```{r}
ggplot() +
  geom_sf(data = md_ev_map, aes(fill = EV_Pct_Change)) +
  theme_minimal()
```
```{r}
#census_api_key("29307c4d6092f28ad0c5fc9ecc4e26f0d1f7811aE", install=TRUE)
```

```{r}
md_county_population <- get_acs(
  geography = "county",
  variables = c(population = "B01001_001"),
  year = 2021,
  state = "MD",
  geometry = TRUE
) |>
  rename(County = NAME) |>
  mutate(
    County = str_to_title(str_remove_all(County, ", Maryland"))
  ) |>
  rename(population = estimate) |> 
  select(County, population, geometry)
```

```{r}
md_county_population <- md_county_population |>
  mutate(County = str_remove(County, " County$"))
```

```{r}
ev_with_pop <- md_county_population |>
  left_join(ev_data, by = "County") |>
  mutate(EV_per_capita = (march_2023 / population) * 1000)
```

```{r}
ggplot() +
  geom_sf(data = ev_with_pop, aes(fill = EV_per_capita)) +
  theme_minimal() +
  scale_fill_viridis_c(option = "magma")
```
